---
page-layout: full
title-block-banner: true
---


### [Learning Agenda]{style="background: lightblue; color: color"} 


1. [An introduction](https://raw.githack.com/lugoga/kitaa/main/slides/1.prerequisite.html#/key-concepts)
1. [Get familiar with R and Rstudio](https://raw.githack.com/lugoga/kitaa/main/slides/2.%20installs.html#/title-slide)
1. [Data types](https://raw.githack.com/lugoga/kitaa/main/slides/4.data_types.html#/title-slide)
1. [Data structures](https://raw.githubusercontent.com/lugoga/kitaa/main/slides/5.data_structure.html)
1. [Reading and writing data in Rstudio](https://raw.githack.com/lugoga/kitaa/main/slides/6.xlds_csv_format.html)
1. [Tidying Data with tidyverse](https://raw.githack.com/lugoga/kitaa/main/slides/7.%20data_manipulation.html#/title-slide)
1. Plotting
      1. [Plotting and Visualization_A](https://raw.githack.com/lugoga/kitaa/main/slides/7_data_visualization.html#/title-slide)
      1. [Plotting and Visualization_B](https://raw.githack.com/lugoga/kitaa/main/slides/8_data_visualization.html)
      1. [Plotting and Visualization_C](https://raw.githack.com/lugoga/kitaa/main/slides/9_ggplot2_continues.html)
1. Data Manipulation
1. Descriptive Statistics
1. Inferential Statistics
1. Modelling and simulation
1. Spatial Handling and Analysis
1. Further topics
  
    1. GIt and Github
    1. Reproducibility with Quarto
    1. Websites and blog
    1. Using python from Rstudio
    1. Generating HTML, PDF and Word Reports
    
    
### [Datasets]{style="background: lightblue; color: color"} 

1. **Chinook** -- contains the lengths and weights for Chinook Salmon from three locations in Argentina. you can simply download the file from the internet using the code chunk below; 

```{r}
#| eval: true
#| 

chinook = readr::read_csv(
  file = "https://raw.githubusercontent.com/lugoga/kitaa/main/datasets/chinook_lw.csv")


```

```{r}
chinook |> 
  FSA::headtail() |> 
  flextable::flextable()
```


1. **Mauna Loa Atmospheric CO2 Concentration**-- store Atmospheric concentrations of CO2 are expressed in parts per million (ppm) and reported in the preliminary 1997 SIO manometric mole fraction scale. The link to download is found in the chunk


```{r}
co = readr::read_csv("https://raw.githubusercontent.com/lugoga/kitaa/main/datasets/long_form_co2.csv")
```

`

```{r}
co |> 
  FSA::headtail() |> 
  flextable::flextable()
```
